---
title: "Aapa distribution map of Norway"
format: pdf
author: Anders Kolstad
date: last-modified
---

This script creates a distribution map of appa mires in Norway, for use in reporting for the Bern convention.
The resolution is 10x10 km.
The distribution is defined based on certain bioclimatic sones and sections.


```{r setup}
#| warning: false
#| message: faalse
library(tidyverse)
#library(dbConnect)
library(dbplyr)
library(DBI)
library(RPostgres)
library(stars)
library(sf)
library(terra)
library(tmap)
```


Connecting to the database where the bioclimatic sones and sections dataset is
```{r connectionObject}
# the password is stored in a pgpass.conf file
con <- DBI::dbConnect(drv = RPostgres::Postgres(), host = "t2lippgsql03", dbname = "ano_moduler")
```


Bring in the data
```{r getData}
bc <- dplyr::tbl(con, dbplyr::in_schema("helper_variables", "bioclimatic_regions"))

bc <- bc |>
  mutate(geom_wkb = dbplyr::sql("ST_AsBinary(geom)")) |>
  select(-geom) |>
  collect() |>
  mutate(geom = sf::st_as_sfc(geom_wkb, crs = 25833)) |>
  sf::st_as_sf(sf_column_name = "geom")

```


Define distribution
```{r define}
reg <- unique(bc$BCregion)
# we exclude boreonemoral and south boreal sones (6SO1-2 and 6SE1)
exl <- grep("6SO-1-2|6SE-1", reg)

# 1 = presence; 0 = absence
bc2 <- bc |>
  mutate(aapa = case_when(
    BCregion %in% reg[exl] ~ 0,
    .default = 1
  ))

# test:
bc2 |> as_tibble() |> count(aapa)
# OK
```

A visual test.
```{r}
bc2 |>
  slice_sample(prop = 0.2) |>
  tmap::tm_shape() +
  tmap::tm_fill(col = "aapa", style = "cat")
```

Checking/confirming that the area of the square polygons are 1 km2
```{r checkResolution}
temp <- bc2 |>
    mutate(km2 = units::drop_units(st_area(geom))*10^-6)

summary(temp |> as_tibble() |> pull(km2))
# area is 1 km1
```

```{r sumfunc}
# defining a summary function
dplyr_summary <- function(data, var) {
  if (inherits(data, "sf")) {
    data <- sf::st_drop_geometry(data)
  }
  data %>%
    summarise(
      Min    = min({{ var }}, na.rm = TRUE),
      Q1     = quantile({{ var }}, 0.25, na.rm = TRUE),
      Median = median({{ var }}, na.rm = TRUE),
      Mean   = mean({{ var }}, na.rm = TRUE),
      Q3     = quantile({{ var }}, 0.75, na.rm = TRUE),
      Max    = max({{ var }}, na.rm = TRUE)
    )
}

```

Import the EEA 10km standard grid
```{r importGrid}
grid <- readRDS(here::here("data", "grid50_kyst_bgr.rds"))

# Test that area is 10 km2
grid |>
  mutate(km2 = units::drop_units(st_area(geometry))*10^-6) |>
  dplyr_summary(km2)

# the area varies a bit.
```

Creating a regular raster aligned with `grid`.

```{r rasterizeGrid}
# set resolution
rr <- 10000
# Get bbox
bb <- st_bbox(grid) 
st_crs(grid)

# create regular grid frm bbox
gr0 <- rast(xmin= floor(bb["xmin"]/rr)*rr, xmax= ceiling(bb["xmax"]/rr)*rr, 
              ymin= floor(bb["ymin"]/rr)*rr, ymax= ceiling(bb["ymax"]/rr)*rr, 
              res= rr, crs= "EPSG:25833")
ret <- rasterize(grid, gr0, cover=T, touches=T) # (for very small polygons touches=TRUE is needed, otherwise FALSE)

class(ret)
```

Rasterize and aggregate the 1km2 polygon map
```{r rasterize}
bc_raster_grid <- rasterize(bc2, ret, fun = "max", field = "aapa") 
res(bc_raster_grid) 
# OK
```

Import outline of Norway and overlay in the same map to check that the CRS is OK
```{r}
nor <- readRDS(here::here("data", "norway.rds"))
class(nor)
st_crs(nor)
```
```{r}
tm_shape(bc_raster_grid)+
  tm_raster("aapa") +
  tm_shape(nor) +
  tm_borders()
# OK
```

```{r}
summary(bc_raster_grid)
```

```{r}
ncell(bc_raster_grid)
```

There are quite a few NA's, but these are from the areas outside mainland Norway (the ocean, Sweden etc.). So that's OK.

I will export in two CRS's.

```{r transform}
bc_raster_3035 <- project(bc_raster_grid, "EPSG:3035", method = "near")
```

Exporting the map
```{r export}
#| eval: false
writeRaster(bc_raster_grid, "output/aapa_dist_map_25833.tif", overwrite=TRUE)
writeRaster(bc_raster_3035, "output/aapa_dist_map_3035.tif", overwrite=TRUE)
```